---
title: "A Gentle Introduction to Causal Inference"
subtitle: "CDCS"
output: html_document
---

Welcome to the R tutorial for the Gentle Introduction to Causal Inference course with the CDCS. In this script you will use your R skills to employ causal inference in practice. For this we will be using the 'palmer penguins' dataset. More information can be found about this data set here: https://allisonhorst.github.io/palmerpenguins/index.html. For time purposes we will use the csv version of this file from the github.

### Exercise 0
Throughout remember we will need tidyverse, go ahead and do this as your first task.

Your Answer:
```{r Exercise0}
library()
```

-------------

### Exercise 1
Go ahead and load in the data below. For this we use the tidyverse function 'read_csv()'.

```{r Exercise1}
PenguinData <- read_csv()
```

-------------

### Exercise 2
Use your favorite functions to explore the data. Write a small summary of the data, include what variables are involved.

```{r Exercise 2}

```

-------------

### Exercise 3
Start to consider what I may want as my exposure (independent variable) and what would be my outcome (dependent variable). Which variables may be confounders and which may be mediators? Come up with **two** different set-ups which would follow the causal inference framework and list them below. Additionally if you have some pen and paper nearby, draw the diagram to directed acyclic graph (DAG) which goes with them.

INSERT YOUR ANSWER HERE 

-------------

### Exercise 4
Lets try and work out the average causal effect of 'sex' on 'flipper_length_mm'. Firstly we will do this manually. To do this you need to calculate the average of the flipper lengths for the males and then for the females. We then take these quantities away from each. You may need to filter the N/A values out! What is the interpretation of this?


```{r Exercise 4}

```

-------------

### Exercise 5
Once again lets calculate the average causal effect of 'sex' on 'flipper_length_mm'. This time, use a linear model and obtain the correct coefficient.

```{r, Exercise5}

```

### Exercise 6
Now that you have worked out the ATE for these basic models, try to implement the models that you worked out in Exercise 3. My recommendation is that you do this via the linear model.

```{r, Exercise 6}

```